### 2009-08-28 paper/2009-09/compare.tex
Firming up quantification section

This list of questions is a modified version of the one presented in \cite{b:compare-ada-c-pascal}.
It is presented here as an organizational aid;
when describing the PPS they have used,
implementors are encouraged to use the order and terminology of this list where appropriate.
We acknowledge that many of these questions will not apply to particular systems,
and encourage implementors to ask questions which we have left out.

>>>>

This list of questions is a modified version of the one presented in \cite{b:compare-ada-c-pascal}.
It is presented here as an organizational aid;
when describing the programming system they have used,
implementors are encouraged to use the order and terminology of this list where appropriate.
We acknowledge that many of these questions will not apply to particular systems,
and encourage implementors to ask questions which we have left out.

### 2009-08-28 paper/2009-09/quant.tex
Rewriting the introduction
Firming up quantification section

\section{Quantifying the Importance of Usability\label{s:corollary}}

According to Amdahl's Law,
the time $t$ required to execute a program has two components:
$t_{seq}$, representing sequential operations that cannot be parallelized,
and $t_{par}$, representing those operations which can.
If we let $\sigma$ be the ``serial fraction'' of the program,
so that $t_{seq} = {\sigma}t$ and $t_{par} = (1-{\sigma})t$,
then the speedup which can be achieved using $P$ processors is:
\begin{eqnarray*}
s(P)	& =	& \frac{t(1)}{t(P)}	\\
	& =	& \frac{t}{{{\sigma}t} + {{(1-{\sigma})t}}/{P}}
\end{eqnarray*}
As $P{\rightarrow}{\infty}$,
the potential speedup is bounded by $1/{\sigma}$.

What is not usually acknowledged is
the influence of programming system usability on
the potential speedup of sequential programs.
As stated above, Amdahl's Law applies only to a single run of a program.
In the real world, programs are first developed, and then run many times.
If we let $T$ be the total running time of a program over its lifetime,
$D_{seq}$ be the time required to develop the sequential components of a program,
and $D_{par}$ be the time required to develop the parallel components
(or parallelize the sequential components),
then the lifetime speedup we can achieve is given by:
\begin{eqnarray*}
S(P)	& =	& \frac{D_{seq} + T(1)}{D_{seq} + D_{par} + T(P)}	\\
	& =	& \frac{D_{seq} + T}{D_{seq} + D_{par} + {{\sigma}T} + {{(1-{\sigma})T}}/{P}}
\end{eqnarray*}
Assume that we are parallelizing a ``dusty deck'',
i.e.\ that the time $D_{seq}$ has already been invested.
If we let $\phi = D_{par}/T$ be the ratio of parallelization time to total program runtime,
then achievable speedup is limited by:
\begin{eqnarray*}
S_{legacy}(P)	& =	& \frac{T}{D_{par} + {\sigma}T}		\\
		& =	& \frac{1}{{\phi} + {\sigma}}
\end{eqnarray*}
The time required to parallelize a program can therefore be seen
as increasing that program's effective serial fraction.
Unless parallelization time can be substantially reduced,
massively-parallel computing is likely to remain attractive only in those cases in which:
\begin{itemize}
\item	the application is trivially parallel (e.g.\ task-farming a rendering calculation);
\item	the expected total program runtime is very large
	(i.e.\ the program is a package used extensively in a field
	such as computational chemistry or vehicle crash simulation);
	or
\item	cost is not an obstacle (i.e.\ the customer is the NSA).
\end{itemize}

@@

In a 1981 survey article \cite{b:psych-prog-survey}, B.~A.\ Sheil observed that:
\begin{quote}
As practiced by computer science,
the study of programming is an unholy mixture of mathematics,
literary criticism,
and folklore.
\end{quote}
Little has changed since.
Computer scientists pay lip service to the importance of usability,
but are strangely reluctant to try to quantify it,
particularly with respect to the tools they use themselves.
Religious debates about the ``right'' first-year teaching language
or the ``best'' GUI toolkit are common;
systematic use of the experimental method is not.

Most of the experiments which have been done have concentrated on ``programming in the small'' issues,
such as readability and code-pattern recognition
\cite{b:reverse-parsing-2,b:cloze-program-1,b:prog-comprehension-1}.
The little work that has been done on ``programming in the large'' has been anecdotal,
such as \cite{b:compare-ada-c-pascal}.
This book presented various authors' impressions of Pascal, C, and Ada
(before any complete Ada implementations existed).
Its editors formulated a list of questions to use in comparing languages,
but did not use it themselves.
A modified version of this questionnaire,
tailored for parallel programming languages,
is given in Appendix~\ref{s:questions},
and will be used as a guideline in our work.

Two comparative efforts in parallel programming are \cite{b:babb-cases} and \cite{b:salishan}.
The first presented implementations of a simple numerical quadrature program
in more than a dozen different parallel languages
in use on mid-1980s hardware.
The second presented implementations of
the Salishan Problems---Hamming number generation,
isomer enumeration,
skyline matrix reduction,
and a simple discrete event simulator---in
C* \cite{b:dataparallel-c}, Occam~\cite{b:occam}, Ada~\cite{b:ada},
and a variety of dataflow and functional languages.
Both of these books convey the flavor of the languages they describe,
but neither made any effort to compare languages or problem implementations directly.

Inspired by the Salishan Problems,
the first author developed a suite of seven problems \cite{b:cowichan-1}
in order to test the limits of parallel programming systems.
Individual problems were badly load-balanced,
manipulated irregular pointer-based structures,
and so on.
Our experience implementing them was extremely valuable,
as it uncovered many bugs in the programming system used,
and revealed several shortcomings of the system's design.
However,
we also discovered that implementing each problem takes six to eight weeks,
which we feel is prohibitive at this exploratory stage.

>>>>

\section{Quantifying the Importance of Usability\label{s:quant}}

According to Amdahl's Law,
if $t(n)$ is an algorithm's running time on $n$ processors
and $\sigma$ is the algorithm's serial fraction
(i.e., the fraction of the algorithm that cannot be parallelized),
then:
\begin{eqnarray*}
t(p) & = & {\sigma}t(1) + (1 - {\sigma})t(1)/p
\end{eqnarray*}
The speedup on $p$ processors is therefore:
\begin{eqnarray*}
s(p) & = & t(1)/t(p) \\
     & = & t(1)/({\sigma}t(1) + (1 - {\sigma})t(1)/p) \\
     & = & 1/({\sigma} + (1 - {\sigma})/p)
\end{eqnarray*}
As $p{\leftarrow}{\infty}$, the speedup is bounded by $1/{\sigma}$.
This makes intuitive sense:
if 10\% of an algorithm can't be parallelized,
then even with infinite hardware,
the program can only be made to run 10 times faster.

But software doesn't just happen:
it has to be written and debugged,
and these activities usually aren't made faster by parallel hardware\footnote{In fact,
most programmers believe that doing parallel programming is harder and slower than doing it sequentially.}.
If we let $T$ and $S$ be the equivalents of $t$ and $s$ over the program's whole lifetime,
and $D$ be its total development time,
then:
\begin{eqnarray*}
T(p) & = & {\sigma}T(1) + (1 - {\sigma})t(1)/p + D
\end{eqnarray*}
The achievable \emph{lifetime} speedup is then:
\begin{eqnarray*}
S(\infty) & = & 1/({\sigma} + D/T(1))
\end{eqnarray*}

The ratio of development time to (sequential) running time is therefore as important to speedup
as the serial fraction of the algorithm.
Put another way,
the time required to parallelize a program effectively increases the program's effective serial fraction.
Unless parallelization time can be substantially reduced,
parallel programming will only be cost-effective when:
\begin{itemize}
\item	the application is trivially parallel (e.g., task-farming a rendering calculation);
\item	the expected total program runtime is very large
	(i.e., the program is a package used extensively in a field
	such as computational chemistry or vehicle crash simulation);
	or
\item	cost is not an obstacle (i.e., the customer is Disney or the NSA).
\end{itemize}

The aim of this problem suite is to capture enough information about $D$
to allow competing parallel programming systems to be ranked and compared.

@@

Two comparative efforts in parallel programming are \cite{b:babb-cases} and \cite{b:salishan}.
The first presented implementations of a simple numerical quadrature program
in more than a dozen different parallel languages used on mid-1980s hardware.
The second presented implementations of
the Salishan Problems---Hamming number generation,
isomer enumeration,
skyline matrix reduction,
and a simple discrete event simulator---in
C*~\cite{b:dataparallel-c}, Occam~\cite{b:occam}, Ada~\cite{b:ada},
and several dataflow and functional languages.
Both books conveyed the flavor of the languages they describe,
but neither compared languages or problem implementations directly.

More recently,
Lutz has explored quantitative comparison of the relative productivity
of different programming languages \cite{b:prechelt-languages,b:prechelt-platforms}.
Our long-term aim is to replicate his work in the domain of parallel programming.
